The Name of the Title Is Hope

### 2025-01-26 paper.tex
Edits, Revert RTS to non-raking

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. By allowing a workgroup to access and operate on another tile’s data, \emph{ScalarFallback} targets Coffman et al.'s \emph{mutual exclusion} condition. Instead, each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not blocking, it guarantees the termination of the algorithm because each workgroup can perform the reduction of any blocking tile, breaking the inter-workgroup dependency.

@@

To maximize the efficiency of the fallback operation, we employ three strategies. First, although a lookback is performed by a single thread or subgroup, we want all threads in the workgroup to participate. To facilitate this, we use a workgroup-wide control flag to allow threads participating in the lookback to signal the entire workgroup when a fallback is necessary. Second, we explicitly limit the number of times a workgroup may spin while waiting on a tile. If the maximum spin count is exceeded, the fallback operation is initiated. Although this approach still risks performing redundant fallbacks---since a work tile that is late is indistinguishable from one that is blocking---exposing the maximum spin count as a tunable parameter allows us to reduce the likelihood of false-positive fallbacks. Finally, after completing a fallback, we attempt to insert the value into the spine at the position of the blocking tile, preventing any succeeding workgroups from initiating redundant fallbacks.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Regardless of FPG, due to false positives, any given hardware will have a natural fallback-incidence rate $f$. Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering guarantees that memory accesses to the spine fall within an $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache. As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile. As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.

@@

\subsubsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes. \footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

@@

  \item \textbf{Fallback:}
        \begin{itemize}
          \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state:
                \begin{itemize}
                  \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 4.
                  \item If the blocking tile is the first tile in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
                  \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
                \end{itemize}
        \end{itemize}

@@

\subsubsection{Tile State Implementation}
To maintain a coherent view of the tile's state, two bits are required. In \emph{Decoupled-Lookback}, Merrill and Garland~\cite{Merrill2016} propose two strategies for communicating tile state and 32-bit reduction value: Two separate 32-bit values---one for the tile state and the other for the reduction value---with updates using a \emph{memory fence}. One single 64-bit value, bit-packing the tile state and reduction value, with updates via 64-bit atomics.
In our case, neither are viable options. Due to a lack of support across backend APIs, WebGPU does not furnish developers with a memory model, and currently, 64-bit atomics are not supported. Moreover, 64-bit atomics are not supported across most mobile devices, and thus we do not consider their future support an appropriate option for our implementation. Instead, we use two 32-bit values and split the reduction into upper and lower 16-bit segments, reserving the two most significant bits of each 32-bit value to encode the tile state.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}
To accommodate split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed. Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values. Notably, our approach extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

In the context of fallbacks, the blocking tile can exist in any of three states—\emph{Not Ready}, \emph{Ready}, or \emph{Inclusive}—due to several factors. A false-positive fallback may occur when the blocking tile completes its local or inclusive reduction before the fallback finishes. Additionally, multiple workgroups are allowed to perform fallbacks on the same blocking tile, meaning another workgroup may have already updated the tile state. To handle these scenarios, we require a tile state representation and an accompanying atomic operation that ensures a fallback can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state. This challenge is further compounded by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting our use. Instead, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely insert values.
\thomas{Possibly a blurb here about no volatile, so we use all atomics?}

@@

  \item Clone the Dawn repository:
  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
git clone https://dawn.googlesource.com/dawn
  \end{lstlisting}

>>>>

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction \thomas{There is a better wording here. Something like, "because although workgroups \emph{may} share execution resources, they are ultimately scheduled individually." This sentence is incredibly important.}. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress requirements in \emph{ScalarFallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. By enabling workgroups to operate on data from other tiles, \emph{ScalarFallback} prevents deadlocking by targeting the \emph{mutual exclusion} condition described by Coffman et al. While this may involve performing redundant work when a tile is late rather than genuinely blocking, it ensures the algorithm's termination by allowing any workgroup to complete the reduction of a blocking tile, thereby breaking inter-workgroup dependencies.

@@

To optimize the fallback operation, we employ three strategies. Although a lookback is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback. To achieve this, our first strategy is the use of a workgroup-wide \emph{control flag}, which enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. Second, we explicitly limit the number of spins a workgroup can perform while waiting on a tile. If this maximum spin count is exceeded, the fallback operation is triggered. While this approach still risks performing redundant fallbacks---since a late tile is indistinguishable from one which is genuinely blocking---exposing the maximum spin count as a tunable parameter helps minimize false-positive fallbacks. Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing subsequent workgroups from initiating redundant fallbacks.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. No subgroup scheduler is perfectly fair, and thus any given hardware will have a natural fallback-incidence rate $f$, regardless of FPG, due to \emph{false-positive} fallback operations. Consequently, we expect memory bandwidth traffic to increase (and throughput to decrease) by $O(fn)$ to $O((2 + f)n)$. However, just as serial ordering guarantees that memory accesses to the spine fall within an $o$-width sliding window, memory accesses to the scan input are similarly guaranteed to fall within an $ot$-width sliding window, making them also highly likely to reside within some level of the cache. As we later empirically demonstrate, this cache residency significantly reduces the cost of memory access, such that the fallback operation is almost always bound by the compute required to reduce the blocking tile. As a pure reduction incurs very lower overhead, we observe throughput approaching $O(2n)$ on most devices.

@@

\subsubsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes. \footnote{Since we also require a workgroup-wide pure reduce operation, this can be achieved by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

@@

  \item \textbf{Fallback:}
        \begin{itemize}
          \item The workgroup computes the reduction of the blocking tile by processing the corresponding elements directly.
          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
          \item As this update attempt is performed atomically, the workgroup reads back the blocking tile's immediate state:
                \begin{itemize}
                  \item If the state is \emph{Inclusive}, the fallback value is discarded, and the workgroup accumulates the value before proceeding to step 4.
                  \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and proceeds to step 4.
                  \item Otherwise, the workgroup resumes the lookback, starting with the preceding tile.
                \end{itemize}
        \end{itemize}

@@

\subsubsection{Tile State Implementation}
Because a tile may be in three possible states, two bits are required to record its state alongside the 32 bits needed to store the tile's reduction value. Previous \emph{Chained-Scan} architectures maintained a coherent view of both by using either 64-bit atomics with bit packing or memory fences with 32-bit atomics. However, neither approach is viable for our implementation, as WebGPU lacks a memory model and does not support 64-bit atomics. Additionally, 64-bit atomics are unsupported on most mobile devices, making their future adoption impractical for our needs. Instead, we use two 32-bit values, splitting the reduction into upper and lower 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}
To accommodate the split values, we use a single subgroup for the lookback, leveraging subgroup \emph{ballot} to verify that the split value state is consistent across threads, and subgroup \emph{shuffle} to join and split reductions as needed. Since this method requires one thread per 16 bits, and the minimum subgroup size in the WebGPU specification is 4, it supports both 32-bit and 64-bit values. Notably, this extends support to 64-bit scans, which would not be possible even with 64-bit atomics.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

In the context of fallbacks, we require a tile state representation and an accompanying atomic operation that ensures a fallback update attempt can safely transition a tile's state to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, the blocking tile can exist in any of the three states due to two scenarios: a false-positive fallback may have been initiated on a tile that transitions to \emph{Ready} or \emph{Inclusive} before the fallback completes, or another workgroup may have already successfully completed a fallback update, leaving the tile in \emph{Ready}.Consequently, the update mechanism should only overwrite tiles in the \emph{Not Ready} state. This requirement is further constrained by the fact that in WebGPU, the correctness of atomic \emph{Compare-and-Swap} is not guaranteed across all devices, prohibiting its use. To address this, we encode the tile state within the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This design enables the use of atomic \emph{Maximum}, which inherently respects the state hierarchy, to safely perform updates.

@@

  \item Clone the Dawn repository:
  \begin{lstlisting}[basicstyle=\ttfamily\small, frame=single]
git clone https://dawn.googlesource.com/dawn
RESET HEAD here to a specific commit. . . to lock in?
  \end{lstlisting}
